sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-24 doc/appendix-pcod-VAST.Rmd

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- as.numeric(1)

fit <- fit_model(
  settings = settings,
  Lat_i = pcod[, "lat"],
  Lon_i = pcod[, "lon"],
  t_i = pcod[, "year"],
  b_i = pcod[, "density"],
  a_i = pcod[, "effort"],
  input_grid = input_grid,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

>>>>

# effort is 1 when using CPUE instead of observed weight as the response:
pcod$effort <- 1
pcod <- as.data.frame(pcod) # ensure not a tibble
fit <- fit_model(
  settings = settings,
  Lat_i = pcod[, "lat"],
  Lon_i = pcod[, "lon"],
  t_i = pcod[, "year"],
  b_i = pcod[, "density"],
  a_i = pcod[, "effort"],
  input_grid = input_grid,
  working_dir = paste0(here("doc", "appendix-VAST"), "/")
)
```

### 2021-11-24 doc/paper.Rmd

@@

<!-- User interfaces to classes of models fit by these template systems can greatly  -->

<!-- reduce these challenges for the vast majority of models. -->

<!-- keep the above even briefer? -->

<!-- inlabru is "very good" but not entirely familiar to those used to `glm()`, lme4, glmmTMB, or mgcv; -->

<!-- *TODO: the differences are many and more than can be listed here. Keep brief add Table or discussion later about differences and when to pick one over the other?* -->

@@

<!-- - high-level paragraph or two about what the model includes, referencing Fig 1 -->
<!-- - high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?) -->
<!-- - high-level paragraph about simulation capabilities and + testing performed (reference supplement) -->
<!-- - model selection/cross validation paragraph -->
<!-- - derived quantities paragraph (or just put this in the examples?) -->

<!-- - example 1 + figs -->
<!-- - example 2 + figs -->

<!--     - perhaps a paragraph on how it achieves what Jim outlined @thorson2019 as necessary: -->
<!--         -   "VAST represents four main design principles. Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
<!--         -   Condition on missing covariates (check, random fields) -->
<!--         -   Bridge between univariate and multivariate applications (no! should we? -->

<!-- sdmTMB in its current form achieves 3 of the 4 design principles outlined by @thorson2019 for index standardization:  -->

<!-- (e.g., 'catchability' and 'habitat' covariates in VAST merely depend on the values supplied to `predict()` in sdmTMB). -->

<!-- # Model -->

@@

<!-- [paragraph about spde and gaussian process / Matern covariance here] -->

<!-- Paragraph on model selection here-->

@@

<!-- Paragraph on advanced options here?
Specifically, this could address
- PC-priors -- and other penalized priors
- Non-linear stuff / etc
- Linkages to Rstan
- Simulation
-->

@@

<!--         -   "VAST represents four main design principles. Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include." -->
<!--         -   Area weighting (check, helper functions to sum up weighted grid cells) -->
<!--         -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`) -->
<!--         -   Condition on missing covariates (check, random fields) -->
<!--         -   Bridge between univariate and multivariate applications (no! should we? -->
<!-- However, VAST can fit multivariate spatiotemporal models (REF) and Poisson-link delta model (REF), whereas sdmTMB cannot. -->

@@

<!--
# Tables

1. Comparison to related software?

- VAST
- INLA
- mgcv
- glmmTMB
- glmmfields?
- other stuff on CRAN, spBayes etc.

# Figures

1. Flow diagram of functions? Not needed?
1. Speed with various knots and/or spatial complexity in comparison to INLA + VAST + ...?
1. Example application 1 - SDM, show mesh(?) and/or raw data, show predictions and SD in space,
1. Example application 2 - population index with stitched survey?

Content:

-   outline the general need for this type of modelling

-   outline the need for this type of package to make the models accessible and rapid to iterate on

@@

-   table comparison to related packages

@@

-   how it achieves the steps Jim described [@thorson2019] in what a VAST-related package has to be able to do (and how VAST does it vs. how sdmTMB does it)
    -   "VAST represents four main design principles. Although there are many other implementation details that distinguish VAST from previous index-standardization methods, I believe that these four design principles represent the features that any future replacement should include."
    -   Area weighting (check, helper functions to sum up weighted grid cells)
    -   Distinct catchability and habitat covariates (check, decide what goes into `predict()`)
    -   Condition on missing covariates (check, random fields)
    -   Bridge between univariate and multivariate applications (no! should we?)

>>>>

```{r rnaturalearthhires, include=FALSE, echo=FALSE}
if (!require("rnaturalearthhires", quietly = TRUE)) {
  stop("Please install 'rnaturalearthhires'.\n",
  "`remotes::install_github('ropensci/rnaturalearthhires')`")
}
```

@@

@@

<!-- - high-level paragraph about design philosophy (modular [separate predict, residuals, etc.], familiar/clear naming, expensive computations only if needed, work with regular data frames, ...?) -->

@@

@@

@@

@@

@@

@@
